The Anthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2023-06-07 paper.tex
more paper

%%%%%%%%%%% Put your definitions here
\newcommand{\aw}{AnthroWorks3D}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://inserthere.com/ontology/}

%%%%%%%%%%% End of definitions

@@

\begin{table}[htbp]
  \centering
  \caption{Table Title}
  \begin{tabulary}{\textwidth}{lLLl}
    \toprule
	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	Anatomical entity			&						&Physical anatomical entity fma61775\\
	Anatomical structure		&						&Material anatomical entity	fma67165		&Material object\\							
	Spatial anatomical entity	&						&immaterial anatomical entity fma:fma67112	&Space entity\\
	Anatomical space			&						&											&Space region\\
	Anatomical surface			&Median Plane			&Anatomical surface fma24137				&Surface region\\
	Anatomical line				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line fma9657	&Line region\\
	Anatomical point			&Gonion					&Anatomical point							&Point region\\
	Anatomical plane			&Median Plane			&Anatomical plane fma242982\\
	Anatomical axis		 		&Sagittal Axis\\
	Skeleton					&						&Skeleton fma:fma23875\\
	Bone structure\\
	Bone						&Mandibula				&Bone organ fma:fma5018\\

	Bone part					&Protuberantia mentalis	&Related to Segment of bone organ fma:fma281808, Zone of bone organ fma:fma10483\\
	Bone compound				&Cranium				&Comparable to union of Skeletal system fma:fma23881 and subdivision of skeletal system fma:fma85544\\
	Tooth structure				&Dentes Permanentes		&Permanent teeth (fma:fma75152; TA2ID:913)\\
	Tooth						&Dens caninus			&Tooth fma:fma12516\\
	Tooth part					&Cuspis dentis			&fma:fma56481; TA2ID:925\\
	Phenotype					&Height, Sex			&											&Attributive\\							
	Anatomical Property			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&Attributive\\
	Relative anatomical location&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&Relator\\
\bottomrule
\end{tabulary}
\end{table}

\subsection{Anatomical entity}
\newcommand{\anno}[1]{#1}%until merge
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.

>>>>

\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://annosaxfdm.de/ontology/}%todo: insert gfo prefix
\daurl{fma}{http://purl.org/sig/ont/fma/fma}

@@

\begin{table}[htbp]
  \centering
  \caption{Table Title}
  \begin{tabulary}{\textwidth}{lLLl}
    \toprule
	ANNOdc class					&Examples				&FMA equivalent / ID						&GFO superclass	\\
	\midrule
	\anno{AnatomicalEntity}			&						&Physical anatomical entity \fma{61775}\\
	\anno{AnatomicalStructure}		&						&Material anatomical entity	\fma{67165}		&\gfo{MaterialObject}\\							
	\anno{SpatialAnatomical entity}	&						&immaterial anatomical entity fma:\fma{67112}	&\gfo{SpaceEntity}\\
	\anno{AnatomicalSpace}			&						&											&\gfo{SpaceRegion}\\
	\anno{AnatomicalSurface}			&Median Plane			&Anatomical surface \fma{24137}			&\gfo{SurfaceRegion}\\
	\anno{AnatomicalLine}				&Mandibular Angle (ppam dexter-go dexter-piam dexter), Sagittal Axis&Anatomical line \fma{9657}	&\gfo{LineRegion}\\
	\anno{AnatomicalPoint}			&Gonion					&Anatomical point \fma{9658}							&\gfo{PointRegion}\\
	\anno{AnatomicalPlane}			&Median Plane			&Anatomical plane \fma{242982}\\
	\anno{AnatomicalAxis}		 		&Sagittal Axis\\
	\anno{Skeleton}					&						&Osseous skeleton \fma{23875}\\
	\anno{BoneStructure}\\
	\anno{Bone}						&Mandibula				&Bone organ \fma{5018}\\

	\anno{BonePart}					&Protuberantia mentalis	&Related to Segment of bone organ \fma{281808}, Zone of bone organ \fma{10483}\\
	\anno{BoneCompound}				&Cranium				&Comparable to union of Skeletal system \fma{23881} and subdivision of skeletal system \fma{85544}\\
	\anno{ToothStructure}				&Dentes Permanentes		&Permanent teeth (\fma{75152}\\%; TA2ID:913)\\
	\anno{Tooth}						&Dens caninus			&Tooth \fma{12516}\\
	\anno{ToothPart}					&Cuspis dentis			\\%\fma{56481}; TA2ID:925\\
	\anno{Phenotype}					&Height, Sex			&											&\gfo{Attributive}\\							
	\anno{AnatomicalProperty}			&Shape (Degree of morphological expression)	&Anatomical relation, e.g., Has\_shape relation	&\gfo{Attributive}\\
	\anno{RelativeAnatomicalLocation}&superior, lateral, medial, posterior	&Anatomical qualitative coordinate	&\gfo{Relator}\\
\bottomrule
\end{tabulary}
\end{table}

\subsection{Anatomical entity}
An anatomical entity is either an anatomical (material) structure or a spatial anatomical entity, such as \anno{Cranium} or \anno{FrontalPlane}.
